Penalized regression for left-truncated and right-censored survival data

### 2021-06-24 manuscript.tex
Update results with updated packages

The adjusted models tend to be better calibrated for patients with the lowest survival probabilities but poorly calibrated for patients with higher survival probabilities. One possible reason for this result is that the adjusted models pushed many of the hazard ratios to $1$ and did not adequately separate low and high risk patients. Probability rescaling such as Platt scaling or isotonic regression may be effective on the adjusted probabilities here, which resemble the characteristic sigmoid-shaped calibration curves produced by different learning algorithms, including the lasso; however the implementation is less straightforward in a survival context \cite{goldstein2021xcal}.

@@

In our real-world use case, an analyst could have mistakenly concluded that the C-index in the non-adjusted model was indicative of good model performance when it was in fact driven in part by a high correlation between predictors and left truncation. For instance, in our large $p$ model the hazard ratio for the predictor ``year of diagnosis'' moved from being the largest in absolute value to around 1 after adjusting for left truncation. The strong positive relationship between ``year of diagnosis'' on the hazard in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias. This relationship reversed completely after adjusting for left truncation, and illustrates that biases to coefficient estimates are driven by the association between predictors and left truncation time. In the case of ``year of diagnosis'', this association was negative and biased the hazard ratio upwards. While many of the other hazard ratios in the large $p$ model were pushed towards 1 after left truncation adjustment, the direction of this bias depends on the direction of the association with left truncation time and varies by predictor. In cases where predictors and left truncation time are positively associated, hazard ratios are biased downwards because the predictors are associated with more immortal time. Because the high C-index misleadingly suggested good model performance, we recommend that models should be evaluated using both discrimination and calibration based metrics to facilitate interpretation.

>>>>

The adjusted models tend to be better calibrated for patients with the lowest survival probabilities but poorly calibrated for patients with higher survival probabilities. One possible reason for this result is that the adjusted models tended to push the most hazard ratios toward $1$, resulting in less separation of low and high risk patients. Probability rescaling such as Platt scaling or isotonic regression may be effective on the adjusted probabilities here, which resemble the characteristic sigmoid-shaped calibration curves produced by different learning algorithms, including the lasso; however the implementation is less straightforward in a survival context \cite{goldstein2021xcal}.

@@

In our real-world use case, an analyst could have mistakenly concluded that the C-index in the non-adjusted model was indicative of good model performance when it was in fact driven in part by a high correlation between predictors and left truncation. For instance, in our large $p$ model the hazard ratio for the predictor ``year of diagnosis'' moved from being the largest in absolute value to around 1 after adjusting for left truncation. The strong positive relationship between ``year of diagnosis'' on the hazard in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias. This relationship reversed completely after adjusting for left truncation, and illustrates that biases to coefficient estimates are driven by the association between predictors and left truncation time. In the case of ``year of diagnosis'', this association was negative and biased the hazard ratio upwards. While some of the other hazard ratios in the large $p$ model were pushed towards 1 after left truncation adjustment, the direction of this bias depends on the direction of the association with left truncation time and varies by predictor. In cases where predictors and left truncation time are positively associated, hazard ratios are biased downwards because the predictors are associated with more immortal time. Because the high C-index misleadingly suggested good model performance, we recommend that models should be evaluated using both discrimination and calibration based metrics to facilitate interpretation.
